Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2020-03-16 paper/results.tex
small edits from a big re-read

To explore the relationship between rotation period, effective temperature
(\teff ) and velocity dispersion, we calculated \sigmavb \footnote{\sigmavb
was calculated as 1.5$\times$ the median absolute deviation, to mitigate
sensitivity to outliers.} for groups of stars with similar rotation periods
and temperatures, and presumed similar age.
The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by \sigmavb, where \sigmavb\ was
calculated for groups of stars over a grid in $\log_{10}$(period) and
temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
In the appendix of this paper, we show that this assumption appears valid for
stars with Galactic latitude $<$ 15\degrees.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff\ (this grid causes the quantized
    appearance).
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages because they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr (Solar age)
    in both top and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from][]{berger2020}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' surface rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}

@@

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' surface rotational
evolution of K dwarfs, like that observed in open clusters \citep{curtis2019}
and reproduced by models that vary angular momentum transport between stellar
core and envelope with time and mass \citep{spada2019}.
The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
parker1970, kawaler1988, charbonneau2010}.
This aligns with the current assumptions about stellar spin-down, dynamo
theory, and the gyrochronology paradigm that has been in place for decades
\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
According to the \citet{spada2019} model, the radiative cores and convective
envelopes of stars are decoupled at these young ages, \ie\ transportation of
angular momentum from the surface to the core of the star is reduced, so the
surface slows down due to wind-braking but the core keeps spinning rapidly.
According to the data presented in figure \ref{fig:vplot}, at intermediate
ages, the rotation periods of K dwarfs appear {\it constant} with mass, and at
late ages rotation period {\it increases} with {\it increasing} mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

A period of core-envelope decoupling in the evolution of cool dwarfs has been
explored in theoretical models for decades \citep[\eg][]{endal1981,
macgregor1991, denissenkov2010, gallet2013}.
In such models, the angular momenta of the radiative core and convective
envelope are permitted to evolve separately once a radiative core develops on
the pre-main sequence.
A core-envelope decoupling phase in the early angular momentum history of
stars is required in order to reproduce observations of young clusters and
star forming regions \citep[\eg][]{irwin2007, bouvier2008, denissenkov2010,
spada2011, reiners2012} and has become an established element of theoretical
gyrochronology models.
The \citet{lanzafame2015} gyrochronology model quantified the mass-dependence
of the core-envelope coupling timescale and this model was fit to open-cluster
observations in \citet{spada2019}, including new rotation period measurements
for stars down to M$\sim$0.6M$_\odot$ in the NGC 6811 cluster
\citep{curtis2019}.
A similar mass-dependent core-envelope coupling timescale was also found to
explain the observed lithium depletion trends in open clusters by an
independent study \citep{somers2016}.
Although variable angular momentum transport between the surfaces and cores of
stars has been an essential ingredient of stellar evolution models for
decades, the transport mechanism is still unknown.
Among the proposed mechanisms are magneto-hydrodynamical waves resulting from
various magnetic field geometries, and gravity waves \citep[see,
\eg][]{charbonneau1993, ruediger1996, spruit2002, talon2003, spada2010,
brun2011, oglethorpe2013}.

Figure \ref{fig:vplot} reveals another phenomenon of magnetic evolution:
K and M dwarfs remain magnetically active for longer than G dwarfs.
The mass dependence of magnetic activity lifetimes has been demonstrated
previously \citep[\eg][]{west2008, newton2017, kiman2019}.
If the detectability of a rotation period is considered to be a magnetic
activity proxy, then our results provide further evidence for a mass-dependent
activity lifetime.
Figure \ref{fig:vplot} shows that the groups of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that the oldest stars with detectable rotation periods, are
cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
massive stars.
To investigate this idea further, we compared the velocity dispersions of
stars with measured rotation periods, to the velocity dispersions of the
overall \kepler\ sample.
We calculated the velocity dispersions for all stars in the \kepler\ field,
after removing visual binaries, subgiants, stars fainter than 16th magnitude,
and high Galactic latitude stars, following the method described in section
\ref{sec:method}.
We then compared these velocity dispersions to the velocity dispersions of
stars in the \mct\ sample.
If the rotation periods of G stars are only detectable when they are young,
G stars with measured periods should have smaller velocity dispersions than
the overall \kepler\ field.
We calculated the ratio of \sigmavb\ for the entire \Kepler\ sample to
\sigmavb\ for stars with rotation periods published in \mct, as a function of
\teff\ (see figure \ref{fig:compare}).
A larger ratio means the rotating star sample is {\it younger}, on average,
than the overall \Kepler\ population, and a ratio of 1 means that the rotating
stars have the {\it same} age distribution as the overall Kepler sample.
Figure \ref{fig:compare} shows that this ratio is largest for G stars and
approaches unity for K and early M dwarfs.
This indicates that the G stars with detectable rotation periods are, on
average, {\it younger} than the total population of G stars in the Kepler
field.
On the other hand, the late K and early M dwarfs with detectable rotation
periods have a similar age distribution to the overall Kepler population which
suggests that the oldest K and M dwarfs are represented in the \mct\ sample.
This result bolsters the evidence that M dwarf rotation periods are measurable
at older ages than G dwarf rotation periods.
G stars become magnetically inactive and have fewer active surface regions
{\it at a younger age than M dwarfs}.

@@

There is a sharp gap in the population of rotation periods (often called the
\kepler\ period gap), which lies just above the 1 Gyr gyrochrone in the upper
panel of figure \ref{fig:vplot}, whose origin is unknown and is the subject of
much speculation \citep{mcquillan2014, davenport2018, reinhold2019}.
This gap was first identified by \mct, and roughly follows a line of constant
gyrochronal age of around 1.1 Gyr \citep[according to the][gyrochronology
relation]{angus2019}.
Several explanations for the gap's origin have been proposed, including a
discontinuous star formation history \citep{mcquillan2013, davenport2017,
davenport2018} and a change in magnetic field structure causing a brief period
where rotational variability is reduced and rotation periods cannot be
measured \citep{reinhold2019}.

@@

An alternate explanation for the gap is that the \mct\ sample contains two
distinct stellar populations: one young and one old.
If so, the kinematic properties of stars above and below the gap are likely to
be distinctly different.
The bottom panel of figure \ref{fig:gap} shows the velocity dispersions of
stars in the \mct\ sample, with stars subdivided into three groups: those that
rotate more quickly than the main rotation period population (green
triangles), those with rotation periods shorter than the gap (orange squares),
and those with rotation periods longer than the gap (blue circles).
Stars were separated into these three groups using \citet{angus2019}
gyrochronology model, according to the scheme shown in the legend.
Only stars cooler than 5000 K are included in the bottom panel in order to
isolate populations above and below the period gap, which only extends up to a
temperature of $\sim$4600 K in our sample, although \citet{davenport2017}
found that the gap extends to temperatures as hot as 6000 K.
In general, velocity dispersion increases with rotation period because both
quantities increase with age.
There is a smooth transition in velocity dispersion between stars with
rotation periods below and above the gap (orange squares to blue circles),
suggesting that these groups are part of the same Galactic population.
Previously, only the overall velocity dispersions of all stars above and below
the gap have been compared, leading to the assumption that these groups belong
to two distinct populations \citep{mcquillan2014}.
The smooth increase in velocity dispersion across the gap shown here suggests
that stars above and below are part of the same stellar population.
However, it does not rule out the possibility that a brief cessation of star
formation in the Solar neighborhood, around one Gyr ago, may have caused this
gap.

>>>>

To explore the relationship between rotation period, effective temperature
(\teff ) and velocity dispersion, we calculated \sigmavb \footnote{\sigmavb\
was calculated as 1.5$\times$ the median absolute deviation of velocities, to
mitigate sensitivity to outliers.} for groups of stars with similar rotation
periods and temperatures, and presumed similar age.
The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by \sigmavb, where \sigmavb\ was
calculated for groups of stars over a grid in $\log_{10}$(period) and
temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
In the appendix of this paper, we show that this assumption appears valid for
stars with Galactic latitude $<$ 15\degrees.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff\ (this grid causes the quantized
    appearance).
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages because they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr (Solar age)
    in both top and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from][]{berger2020}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' surface rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}

@@

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' surface rotational
evolution of K dwarfs, like that observed in open clusters \citep{curtis2019}
and reproduced by models that vary angular momentum transport between stellar
core and envelope with time and mass \citep{spada2019}.

The velocity dispersions of stars in the \mct\ sample, shown in figure
\ref{fig:vplot}, provide the following picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
parker1970, kawaler1988, charbonneau2010}.
This aligns with the current assumptions about stellar spin-down, dynamo
theory, and the gyrochronology paradigm that has been in place for decades
\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
According to the \citet{spada2019} model, the radiative cores and convective
envelopes of stars are decoupled at these young ages, \ie\ transportation of
angular momentum from the surface to the core of the star is reduced, so the
surface slows down due to wind-braking but the core keeps spinning rapidly.
According to the data presented in figure \ref{fig:vplot}, at intermediate
ages, the rotation periods of K dwarfs appear {\it constant} with mass, and at
late ages rotation period {\it increases} with {\it increasing} mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

A period of core-envelope decoupling in the evolution of cool dwarfs has been
explored in theoretical models for decades \citep[\eg][]{endal1981,
macgregor1991, denissenkov2010, gallet2013}.
In such models, the angular momenta of the radiative core and convective
envelope are permitted to evolve separately once a radiative core develops on
the pre-main sequence.
A decoupled core and envelope is required to reproduce observations of young
clusters and star forming regions \citep[\eg][]{irwin2007, bouvier2008,
denissenkov2010, spada2011, reiners2012} and has become an established element
of theoretical gyrochronology models.
During this phase, angular momentum transport between radiative core and
convective envelope is reduced.
Over time, models increase the efficiency of angular momentum transport
between the core and envelope in order to reproduce the close-to solid body
rotation observed for the Sun \citep[\eg][]{thompson1996}.
The core-envelope coupling timescale affects the predicted surface rotation
periods of young and intermediate-age stars and is usually constrained using
observations of open clusters.
The \citet{lanzafame2015} gyrochronology model uses a mass-dependent
core-envelope coupling timescale, and \citet{spada2019} fit this model to
open-cluster observations, including new rotation period measurements for
K dwarfs in the NGC 6811 cluster \citep{curtis2019}.
A similar mass-dependent core-envelope coupling timescale was also found to
explain the observed lithium depletion trends in open clusters by an
independent study \citep{somers2016}.
Although variable angular momentum transport between the surfaces and cores of
stars has been an essential ingredient of stellar evolution models for
decades, the transport mechanism is still unknown.
Among the proposed mechanisms are magneto-hydrodynamical waves resulting from
various magnetic field geometries, and gravity waves \citep[see,
\eg][]{charbonneau1993, ruediger1996, spruit2002, talon2003, spada2010,
brun2011, oglethorpe2013}.
% These different processes may make subtly different predictions for the
% rotational evolution of stars, and could potentially be tested with the large
% number of field stars with measured ages and rotation periods which could be
% provided by kinematic ages for the \mct\ sample.

Figure \ref{fig:vplot} reveals another phenomenon of magnetic evolution:
K and M dwarfs remain magnetically active for longer than G dwarfs.
The mass dependence of magnetic activity lifetimes has been demonstrated
previously \citep[\eg][]{west2008, newton2017, kiman2019}, and if the
detectability of a rotation period is considered to be a magnetic activity
proxy, then our results provide further evidence for a mass-dependent activity
lifetime.
Figure \ref{fig:vplot} shows that the groups of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that the oldest stars with detectable rotation periods, are
cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
massive stars.
To investigate this idea further, we compared the velocity dispersions of
stars with measured rotation periods, to the velocity dispersions of the
overall \kepler\ sample.
We calculated the velocity dispersions for all stars in the \kepler\ field,
after removing visual binaries, subgiants, stars fainter than 16th magnitude,
and high Galactic latitude stars, following the method described in section
\ref{sec:method}.
We then compared these velocity dispersions to the velocity dispersions of
stars in the \mct\ sample.
If the rotation periods of G stars are only detectable when they are young, G
stars with measured periods should have smaller velocity dispersions than G
stars in the overall \kepler\ field.
We calculated the ratio of \sigmavb\ for the entire \Kepler\ sample to
\sigmavb\ for stars with rotation periods published in \mct, as a function of
\teff\ (see figure \ref{fig:compare}).
A larger ratio means the rotating star sample is {\it younger}, on average,
than the overall \Kepler\ population, and a ratio of 1 means that the rotating
stars have the {\it same} age distribution as the overall Kepler sample.
Figure \ref{fig:compare} shows that this ratio is largest for G stars and
approaches unity for K and early M dwarfs.
This indicates that the G stars with detectable rotation periods are, on
average, {\it younger} than the total population of G stars in the Kepler
field.
On the other hand, the late K and early M dwarfs with detectable rotation
periods have a similar age distribution to the overall Kepler population which
suggests that the oldest K and M dwarfs are represented in the \mct\ sample.
This result bolsters the evidence that M dwarf rotation periods are measurable
at older ages than G dwarf rotation periods.
In other words, G stars become magnetically inactive and have fewer active
surface regions {\it at a younger age than M dwarfs}.

@@

There is a sharp gap in the population of rotation periods (often called the
\kepler\ period gap), which lies just above the 1 Gyr gyrochrone in the upper
panel of figure \ref{fig:vplot}, whose origin is unknown and is the subject of
much speculation \citep{mcquillan2014, davenport2017, davenport2018,
reinhold2019, reinhold2020}.
This gap was first identified by \mct, and roughly follows a line of constant
gyrochronal age of around 1.1 Gyr \citep[according to the][gyrochronology
relation]{angus2019}.
Several explanations for the gap's origin have been proposed, including a
discontinuous star formation history \citep{mcquillan2013, davenport2017,
davenport2018} and a change in magnetic field structure causing a brief period
where rotational variability is reduced and rotation periods cannot be
measured \citep{reinhold2019, reinhold2020}.

@@

An alternate explanation for the gap is that the \mct\ sample contains two
distinct stellar populations: one young and one old.
If so, the kinematic properties of stars above and below the gap are likely to
be distinctly different.
The bottom panel of figure \ref{fig:gap} shows the velocity dispersions of
stars in the \mct\ sample, with stars subdivided into three groups: those that
rotate more quickly than the main rotation period population (green
triangles), those with rotation periods shorter than the gap (orange squares),
and those with rotation periods longer than the gap (blue circles).
Stars were separated into these three groups using \citet{angus2019}
gyrochronology model, according to the scheme shown in the legend.
Only stars cooler than 5000 K are included in the bottom panel in order to
isolate populations above and below the period gap, which only extends up to a
temperature of $\sim$4600 K in our sample, although \citet{davenport2017}
found that the gap extends to temperatures as hot as 6000 K.
In general, velocity dispersion increases with rotation period because both
quantities increase with age.
Previously, only the overall velocity dispersions of all stars above and below
the gap have been compared, leading to the assumption that these groups belong
to two distinct populations \citep{mcquillan2014}.
However, figure \ref{fig:gap} shows a smooth increase in velocity dispersion
with rotation period across the gap (from orange squares to blue circles),
suggesting that these groups are part of the same Galactic population.
% The smooth increase in velocity dispersion across the gap shown here suggests
% that stars above and below are part of the same stellar population.
This observation does not rule out the possibility that a brief cessation of
star formation in the Solar neighborhood, around one Gyr ago, may have caused
this gap, however.
